---
title: "Activity Monitoring"
author: "Carlos Ruiz"
date: "Friday, November 14, 2014"
output: html_document
---

Load the ggplot2 library

```{r}
library("ggplot2")
```

Set locale to avoid different month and weekday names

```{r}
Sys.setlocale("LC_TIME", "English")
```

##Loading and preprocessing the data

###Loading data

Read the **activity.csv** file with the activity data. We assume that the file resides in the current working directory. The default parameters are just fine

```{r}
activity.data <- read.csv("activity.csv")
```

###Transform data

We need to transform the **date** column to **Date** type

```{r}
activity.data[, "date"] <- as.Date(activity.data[, "date"], format = "%Y-%m-%d")
```

##Questions

###Mean total number of steps taken per day

Total number of steps taken each day

```{r, fig.width=10}
# group by date and report the sum of steps taken per day
sum.steps.by.date <- aggregate(steps ~ date, data = activity.data, sum)

# plot the sum
with (sum.steps.by.date, {
    qplot(
       date, 
       steps,
       main = "Sum of steps taken per day",
       xlab = "Date", 
       ylab = "Steps",
       geom = "line"
    )
})

# print the aggregated data
sum.steps.by.date
```

The following is the mean of the total number of steps

```{r}
mean.steps <- mean(sum.steps.by.date$steps)
mean.steps
```

The following is the median of the total number of steps

```{r}
median.steps <- median(sum.steps.by.date$steps)
median.steps
```

Mean and median of the total number of steps taken per day

```{r, fig.width=10}
# group by date and report the mean of steps taken per day
mean.steps.by.date <- aggregate(steps ~ date, data = sum.steps.by.date, mean)

# plot the mean
with (mean.steps.by.date, {
    qplot(
        date, 
        steps, 
        main = "Mean of total steps taken per day",
        xlab = "Date", 
        ylab = "Steps", 
        geom = "line"
    )
})

# print the aggregated data
mean.steps.by.date

# group by date and report the median of steps taken per day
median.steps.by.date <- aggregate(steps ~ date, data = sum.steps.by.date, median)

# plot the median
with (median.steps.by.date, {
    qplot(
        date, 
        steps, 
        main = "Median of total steps taken per day", 
        xlab = "Date", 
        ylab = "Steps",
        geom = "line"
    )
})

# print the aggregated data
median.steps.by.date
```

###Average daily activity pattern

Time series plot of the 5-minute interval and the average number of steps taken 
averaged across all days

```{r, fig.width=10}
# group by interval and report the mean of steps taken per day
mean.steps.by.interval <- aggregate(steps ~ interval, data = activity.data, mean)

# plot the mean
with (mean.steps.by.interval, {
    qplot(
        interval, 
        steps, 
        main = "Mean of steps taken per 5-minute interval",
        xlab = "5-minute interval",
        ylab = "Steps",
        geom = "line"
    )
})

# print the aggregated data
mean.steps.by.interval
```

The following 5-minute interval on average accross all the days in the dataset contains the maximum number of steps

```{r}
mean.steps.by.interval[mean.steps.by.interval[, "steps"] == max(mean.steps.by.interval[, "steps"]), ]
```

###Imputing missing values

Total number of missing values in the dataset

```{r}
sum(is.na(activity.data[, "steps"]))
```

The **strategy** choosen to fill the NA steps values will be the **mean of each 5-minute interval** for the same 5-minute interval

```{r}
# merge by 5-minute interval to be able to replace the NA values with the mean
filled.activity.data <- merge(
    activity.data, 
    mean.steps.by.interval, 
    by.x = "interval", 
    by.y = "interval"
)

# assign the means of each 5-minute interval where steps value are NA
filled.activity.data[is.na(filled.activity.data[, "steps.x"]), "steps.x"] <- 
    filled.activity.data[is.na(filled.activity.data[, "steps.x"]), "steps.y"]
```

Total number of steps taken each day when filling NA values

```{r, fig.width=10}
# group by date and report the sum of steps taken per day
filled.sum.steps.by.date <- aggregate(steps.x ~ date, data = filled.activity.data, sum)

# plot the sum
with (filled.sum.steps.by.date, {
    qplot(
        date, 
        steps.x, 
        main = "Sum of steps taken per day with filled NA values",
        xlab = "Date", 
        ylab = "Steps",
        geom = "line"
    )
})

# print the aggregated data
filled.sum.steps.by.date
```

The following is the mean of the total number of steps when filling NA values

```{r}
filled.mean.steps <- mean(filled.sum.steps.by.date$steps)
filled.mean.steps
```

The following is the median of the total number of steps when filling NA values

```{r}
filled.median.steps <- median(filled.sum.steps.by.date$steps)
filled.median.steps
```

Mean and median of the total number of steps taken per day when filling NA values

```{r, fig.width=10}
# group by date and report the mean of steps taken per day
filled.mean.steps.by.date <- aggregate(steps.x ~ date, data = filled.sum.steps.by.date, mean)

# plot the mean
with (filled.mean.steps.by.date, {
    qplot(
        date, 
        steps.x, 
        main = "Mean of total steps taken per day with filled NA values",
        xlab = "Date",
        ylab = "Steps",
        geom = "line"
    )
})

# print the aggregated data
filled.mean.steps.by.date

# group by date and report the median of steps taken per day
filled.median.steps.by.date <- aggregate(steps.x ~ date, data = filled.sum.steps.by.date, median)

# plot the median
with (filled.median.steps.by.date, {
    qplot(
        date, 
        steps.x, 
        main = "Median of total steps taken per day with filled NA values", 
        xlab = "Date", 
        ylab = "Steps", 
        geom = "line"
    )
})

# print the aggregated data
filled.median.steps.by.date
```

There are differences from the estimates when ignoring NA values. For example, the first day (2010-10-01) was not plotted because all the steps values were NA

Comparing the original mean and median against the mean and median with filled NA values

```{r}
cat(mean.steps, filled.mean.steps)
```

```{r}
cat(median.steps, filled.median.steps)
```

There is a slightly difference between the medians. When filling the NA values with the mean of the same 5-second interval, the mean and the median are the same (symmetric distribution)

###Differences in activity patterns between weekdays and weekends

```{r, fig.width=10 }
# get the weekday from date column
filled.activity.data[, "day"] <- weekdays(filled.activity.data[, "date"], TRUE)

# assign weekends
filled.activity.data[filled.activity.data[, "day"] %in% c("Sat", "Sun"), "day"] <- "weekend"

# assign weekdays
filled.activity.data[filled.activity.data[, "day"] != "weekend", "day"] <- "weekday"

# group by interval and day
filled.mean.steps.by.interval <- aggregate(steps.x ~ interval + day, data = filled.activity.data, mean)

# plot the mean
with (filled.mean.steps.by.interval, {
    qplot(
        interval, 
        steps.x, 
        facet = day, 
        main = "Differences in activity patterns between weekdays and weekends", 
        xlab = "5-minute interval", 
        ylab = "Mean of steps", 
        geom = "line", 
        color = day
    )
})

# print the aggregated data
filled.mean.steps.by.interval
```

There are some differences in several 5-minute intervals when the day are a weekend day. I decided to plot both the weekdays and the weekends in the same plot to be able to compare easily